Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-18 introduction.tex
Adding Horizon computation.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1110 ccdd05d5-56c1-461a-bf98-85f6876523b4
Adjusting Table 1 location.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1111 ccdd05d5-56c1-461a-bf98-85f6876523b4
lloid_paper: Telescopic -> EM
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1114 ccdd05d5-56c1-461a-bf98-85f6876523b4
Adding localization figure
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1118 ccdd05d5-56c1-461a-bf98-85f6876523b4

It has been observed that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection. Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable
within 25~s of merger. These rates are shown with their substantial uncertainties in figure~\ref{fig:earlywarning}.
%
\begin{table}
\caption{\label{table:sky-localization-accuracy}Horizon distance, expected detection rate, and \SNR\ at merger for a source that produces an \SNR\ of 8 in all three detectors at time $t$ before merger.  The fifth and sixth columns show the minimum area of the 90\% confidence region before merger and at merger.}
\begin{center}
\begin{tabular}{rrrrrr}
\tableline\tableline
& horiz. & rate & final & \multicolumn{2}{c}{$A$(90\%) (deg$^2$)} \\
\cline{5-6}
$t$ (s) & (Mpc) & (yr$^{-1}$) & \SNR\ & early & final \\
\tableline
25.0 & 203 & 5 & 15.8 & 2893.2 & 1.8 \\
10.0 & 269 & 12 & 11.9 & 1150.6 & 3.6 \\
5.0 & 307 & 18 & 10.4 & 594.6 & 5.1 \\
1.0 & 360 & 29 & 8.9 & 153.9 & 7.5 \\
0.1 & 391 & 37 & 8.2 & 32.5 & 9.1 \\
0.0 & 400 & 40 & 8.0 & 9.6 & 9.6 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

Telescopic followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{table:sky-localization-accuracy} shows the
minimum 90\% confidence area for a source that produces an \SNR\ of 8 in the
\LIGO\ detectors in Hanford, Washington, and Livingston, Louisiana, and, the
Virgo detector in Cascina, Italy, at a time $t$ before merger.  Also shown are
the final \SNR\ and 90\% confidence area at merger for the same source.
At 10~s before merger, the pre-merger 90\% confidence area is about $1150.6$~deg$^2$
(2.8\% of the sky) and at 1~s before merger is 153.9~deg$^2$ (0.37\% of the sky).
The confidence region shrinks quadratically in area as the overall amplitude of the
signal increases.

>>>>

\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
It has been observed that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection. Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable
within 25~s of merger. These rates are shown with their substantial uncertainties in figure~\ref{fig:earlywarning}.

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, expected detection rate, and \SNR\ at merger for sources that produce an \SNR\ of 8 in all three detectors at time $t$ before merger.  The fifth and sixth columns show the minimum area of the 90\% confidence region before merger and at merger.}
\begin{center}
\begin{tabular}{rrrrrr}
\tableline\tableline
& horiz. & rate & final & \multicolumn{2}{c}{$A$(90\%) (deg$^2$)} \\
\cline{5-6}
$t$ (s) & (Mpc) & (yr$^{-1}$) & \SNR\ & early & final \\
\tableline
25.0 & 226 & 5 & 15.8 & 2893.2 & 1.8 \\
10.0 & 299 & 12 & 11.9 & 1150.6 & 3.6 \\
5.0 & 341 & 18 & 10.4 & 594.6 & 5.1 \\
1.0 & 401 & 29 & 8.9 & 153.9 & 7.5 \\
0.1 & 435 & 37 & 8.2 & 32.5 & 9.1 \\
0.0 & 445 & 40 & 8.0 & 9.6 & 9.6 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time.pdf}
\caption{}
\end{figure}
%
\EM{} followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{table:sky-localization-accuracy} shows the
minimum 90\% confidence area for a source that produces an \SNR\ of 8 in the
\LIGO\ detectors in Hanford, Washington, and Livingston, Louisiana, and, the
Virgo detector in Cascina, Italy, at a time $t$ before merger.  Also shown are
the final \SNR\ and 90\% confidence area at merger for the same source.
At 10~s before merger, the pre-merger 90\% confidence area is about $1150.6$~deg$^2$
(2.8\% of the sky) and at 1~s before merger is 153.9~deg$^2$ (0.37\% of the sky).
The confidence region shrinks quadratically in area as the overall amplitude of the
signal increases.
